Synthesising Recursive Functions for First-Order Model Counting:\\ Challenges, Progress, and Conjectures

### 2023-03-01 doc/paper/paper.tex
Various changes to the paper

@@

\crefname{algocf}{Algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{condition}{Condition}{Conditions}
\Crefname{condition}{Condition}{Conditions}

@@

\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{line}{line}{lines}

@@

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. The development of
lifted inference algorithms coincided with work on probabilistic relational
models that combine the syntactic power of first-order logic with probabilistic
models such as Bayesian and Markov networks, allowing for a more relational view
of uncertainty
modelling~\cite{DBLP:series/synthesis/2016Raedt,DBLP:journals/ml/KimmigMG15,DBLP:journals/ml/RichardsonD06}.
Lifted inference techniques for probabilistic databases, while developed
somewhat independently, have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC has received more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21}. There is also a recent line
of work that uses FOMC-based methods to count solutions to constraint
satisfaction problems~\cite{DBLP:journals/jair/TotisDRK23}.

@@

% Magician: an approach to improve the efficiency of sampling and local search
% based MLN inference algorithms by utilising \#CSP algorithms. Tuffy: grounding
% and stochastic local search.

@@

% TODO: must remove this footnote. Also, two more papers to mention:
% 1. the L2C paper has the factorial example as well. But then why doesn't L2C work on it?
% 2. cite the new paper "On Discovering Interesting Combinatorial Integer Sequences" as another example of computing n! using publicly unavailable code and as an example of applications to combinatorics
However, none of the publicly available exact (W)FOMC algorithms can efficiently
compute functions as simple as a factorial.\footnote{There is some previous work
  discussing FOMC for
  factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
  One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
  not directly correspond to any algorithm.
  Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
  liftable but contains no experimental results. The
  third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
  with an algorithm that is not publicly available.} We claim that the
capabilities of (W)FOMC algorithms can be significantly extended by empowering
them with the ability to construct recursive solutions. The topic of recursion
in the context of WFOMC has been studied before but in limited ways.
Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21} use WFOMC to
generate numerical data that is then used to conjecture recurrence relations
that explain that data. Van den Broeck~\shortcite{DBLP:conf/nips/Broeck11}
introduced the idea of \emph{domain recursion}. Intuitively, domain recursion
partitions a domain of size $n$ into a single explicitly named constant and the
remaining domain of size $n-1$. However, many stringent conditions are enforced
to ensure that the search for a tractable solution always terminates.

@@

% TODO: c = (L, E, \delta), E - inequality constraints
\begin{definition}[Clause]\label{def:clause}
  A \emph{clause}
  is a triple $c = (L, C, \delta_c)$, where $L$ is a set of literals, $C$ is a
  set of constraints, and $\delta_c$ is the domain map of $c$. Let $\Vars$ be
  the function that maps clauses and sets of either literals or constraints to
  the set of variables in them. In particular,
  $\Vars(c) \coloneqq \Vars(L) \cup \Vars(C)$. \emph{Domain map}
  $\delta_{c}\colon \Vars(c) \to \mathcal{D}$ is a function that maps all
  variables in $c$ to their domains such that (s.t.) if $(X, Y) \in C$ for some
  variables $X$ and $Y$, then $\delta_c(X) = \delta_c(Y)$. For convenience, we
  sometimes write $\delta_c$ for the domain map of $c$ without unpacking $c$
  into its three constituents.
\end{definition}

@@

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.
First, we decouple variable-to-domain assignments from constraints and move them
to a separate function $\delta_{c}$ in \cref{def:clause}. Second, while they
allow for equality constraints and constraints of the form $X \not\in \Delta$
for some variable $X$ and domain $\Delta$, we exclude such constraints simply
because they are not needed. Moreover, note that Van den Broeck et
al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} refer to clauses and formulas as
c-clauses and c-theories, respectively.
% TODO: explain why we're not losing anything by dropping equality constraints
% (e.g., X = Y). Answer: having two variables that are equal is the same thing
% as having one variable (since by the definition of a clause, this equality can
% only be at the beginning of a clause.

One can read a formula in our format as a sentence in first-order logic. All
variables in a clause are implicitly universally quantified, and all clauses in
a formula are implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{align*}
  &\begin{multlined}[t]
    (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
    Y \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(X, Z)) \land{}
    \end{multlined}\\
  &\begin{multlined}[t]
    (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
    X \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(Z, Y)).
    \end{multlined}
\end{align*}

@@

\section{From Circuits to Graphs}\label{sec:methods}

@@

% TODO: the definition of FCG is hard to miss (highlight its importance, maybe
% remind the reader later on in the paper)

% TODO: instead of FO d-DNNF computational graph, say 'first-order computational
% graph' (FOCG?)

In this paper, we introduce \emph{FO d-DNNF computational graphs} (FCGs) that
generalise circuits by dispensing with acyclicity. Formally, an FCG is a (weakly
connected) directed graph with a single source, node labels, and ordered
outgoing edges. Node labels consist of two parts: the \emph{type} and the
\emph{parameters}. The type of a node determines its out-degree. We make the
following changes and additions to the node types already supported by
\textsc{ForcLift}. First, we introduce a new node type for constraint removal
($\CR$). Second, we replace domain recursion with generalised domain recursion
($\GDR$). And third, for reasoning about partially-constructed FCGs, we write
$\star$ for a placeholder type that is yet to be replaced. We write $T_p$ for an
FCG that has a node with label $T_p$ (i.e., type $T$ and parameter(s) $p$) and
$\star$'s as all of its direct successors. We also write $T_p(v)$ for an FCG
with one edge from a node labelled $T_{p}$ to some other node $v$ (and no other
nodes or edges). See \cref{fig:examplefcg} for an example FCG\@.

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\GDR$ nodes.}\label{alg:domainrecursion}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $\Omega \in \mathcal{D}$ s.t.\ there is $c \in \phi$ and $X \in \Vars(L_c)$ s.t. $\delta_c(X) = \Omega$\label{line:condition}}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant in domain } \Omega$\;\label{line:constant}
    \ForEach{clause $c = (L, C, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, X \in \Vars(L) \mid \delta(X) = \Omega \,\}$\;\label{line:V}
      \ForEach{$W \subseteq V$ s.t. $W^2 \cap C = \emptyset$ {\bf and} $W \cap \{\, X \in \Vars(C) \mid (X, y) \in C \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{$\delta'$ restricts $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], C[x/W] \cup \{\, (X, x) \mid (X \in V \setminus W) \,\}, \delta') \,\}$\;\label{line:generation}
      }
    }
    $S \gets S \cup \{\, (\GDR, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in evidently
  unsatisfiable formulas. We impose two restrictions on $W$. First,
  $W^2 \cap C = \emptyset$ ensures that there are no pairs of variables in $W$
  that are constrained to be distinct, since that would result in an $x \ne x$
  constraint after substitution. Similarly, we want to avoid variables in $W$
  that have inequality constraints with constants: after the substitution, such
  constraints would transform into inequality constraints between two constants.
  In this case, both subsets of $V$ satisfy these conditions, and
  \cref{line:generation} generates two clauses for the output formula:
  \begin{multline*}
    (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg \texttt{p}(x, Y), \neg \texttt{p}(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

@@

Note that \Cref{thm:correctness1} is equivalent to Proposition~3 by Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11}. For the proofs of this and other
theorems, see supplementary material.
% TODO (about the first sentence): say: this shows that there is no loss of
% expressiveness/compilation completeness with our new algorithm. Would be nice
% if you could also show formally how it has improved on Guy, e.g., Prop.
% \exists formula \phi such that AlgoGuy(\phi) may fail but NewAlg(\phi) \ne
% \emptyset. Therefore, our algorithms is `more' complete than Guy's

@@

\begin{definition}\label{def:replaceable}
  With respect to a formula $\phi$, a pair $(\Omega, x)$ of a domain
  $\Omega \in \mathcal{D}$ and its element $x \in \Omega$ is called
  \emph{replaceable} if
  \begin{itemize}
    \item $x$ does not occur in any literal of any clause of $\phi$,
    \item and for each clause $c = (L, C, \delta_c) \in \phi$ and variable
          $X \in \Vars(c)$, either $\delta_c(X) \ne \Omega$ or $(X, x) \in C$.
  \end{itemize}
\end{definition}

\begin{algorithm}[t]
  \caption{The compilation rule for $\CR$ nodes.}\label{alg:constraintremoval}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{replaceable pair $(\Omega \in \mathcal{D}, x \in \Omega)$\label{line:crconditions}}{
    add a new domain $\Omega'$ to $\mathcal{D}$\;\label{line:newdomain}
    $\phi' \gets \emptyset$\;
    \ForEach{clause $(L, C, \delta) \in \phi$}{
      $C' \gets \{\, (a, b) \in C \mid b \ne x \,\}$\;\label{line:constraintremoval}
      \nosemic$\delta' \gets X \mapsto
      \begin{cases}
        \Omega' & \text{if } \delta(X) = \Omega\\
        \delta(X) & \text{otherwise;}
      \end{cases}$\;\label{line:newdelta}
      $\phi' \gets \phi' \cup \{\, (L, C', \delta') \,\}$\;
    }
    $S \gets S \cup \{\, (\CR_{\Omega \mapsto \Omega'}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{theorem}[Correctness of $\CR$]
  Let $\phi$ be the input formula of \cref{alg:constraintremoval}, $(\Omega, x)$
  a replaceable pair, and $\phi'$ the formula constructed by the algorithm, when
  $(\Omega, x)$ is selected on \cref{line:crconditions}. Then
  $\phi \equiv \phi'$, where the domain $\Omega'$ introduced on
  \cref{line:newdomain} is interpreted as $\Omega \setminus \{\, x \,\}$.
\end{theorem}
% TODO: Does this have analogue in Guy? If yes, explain difference. If no,
% emphasize this point. Same for Theorem 3.

@@

\paragraph{Hashing.}
We use (integer-valued) hash functions to discard pairs of formulas that are too
different for recursion to work. The hash code of a clause
$c = (L, C, \delta_{c})$ (denoted by $\# c$) combines the hash codes of the sets
of constants and predicates in $c$, the numbers of positive and negative
literals, the number of inequality constraints $|C|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

@@

@@

\begin{itemize}
  \item TODO: is maybe `parameter name' (or something similar) would be more
        appropriate than `variable name'?
  \item We call a variable name is \emph{atomic} if it is free of subtraction
        (e.g., $m$, $n$); otherwise it is \emph{non-atomic} (e.g., $m-1$,
        $n-l$).
  \item For each node $v$, let $\mathfrak{D}(v)$ denote the (precomputed) set of
        domains, the sizes of which we would use as parameters if we were to
        define a function that begins at $v$. This set is computed by
        iteratively setting
        $\mathfrak{D}(v) \gets \left(\bigcup_{u} \mathfrak{D}(u) \setminus I_{v}\right) \cup U_{v}$
        until convergence, where
        \begin{itemize}
          \item $u$ is any direct successor of $v$,
          \item $I_{v}$ is the set of domains \emph{introduced} at node $v$, and
          \item $U_{v}$ is the set of domains \emph{used} by $v$.
        \end{itemize}
  \item We assume a fixed total ordering of all domains names. In particular, in
        our running example we assume that $\Gamma$ goes before $\Delta$.
  \item $\mathsf{F}$ is a partial map from the set of FCG nodes to function
        names (e.g., $f$, $g$).
  \item \newVariableName just generates a unique identifier for a variable
  \item \newFunctionName does the same thing for function names and also records
        the new name in $\mathsf{F}$
  \item The first function in the list is the main one.
  \item Note that a node is a direct successor of a $\Reff$ node if and only if
        it has an in-degree greater than one.
  \item For simplicity, we refer to FCG nodes using their types.
  \item The algorithm consists of two main functions: \visit and \actuallyVisit.
        The former handles new function definitions, and the latter produces an
        algebraic interpretation of each node depending on its type. As there
        are many node types, \cref{alg:interpretation} only includes the ones
        pertinent to the contributions of this paper; see previous
        work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information about the other
        types.
  \item Given some node $v$ as input, both \visit and \actuallyVisit return a
        pair $(\mathsf{e}, \mathsf{funs})$. Here, $\mathsf{e}$ is the algebraic
        expression that should be used to evaluate $v$ or use it in another
        expression, and $\mathsf{funs}$ is a list of auxiliary functions that
        were created while constructing $\mathsf{e}$.
  \item TODO: $\mathfrak{D}$ must be a list. That's the ordering of domain names
        comes in, not in list comprehensions.
\end{itemize}

@@

However, the most important direction for future work is to fully automate this
new process. First, we need an algorithm that transforms FCGs into definitions
of functions. Formalising this process would also allow us to prove the
correctness of the new compilation rules in constructing FCGs that indeed
compute the right (weighted) model counts. Second, these definitions must be
simplified before they can be used, perhaps by a computer algebra system. Third,
we need a way to find the base cases for the recursive definitions provided by
\textsc{Crane}. Fourth, since the first solution found by \textsc{Crane} is not
always optimal in terms of its complexity, an automated way to determine the
asymptotic complexity of a solution would be helpful as well. Achieving these
goals would enable \textsc{Crane} to automatically constructing efficient ways
to compute various functions and sequences. In addition to the potential impact
on areas of artificial intelligence (AI) such as statistical relational
AI~\cite{DBLP:series/synthesis/2016Raedt}, \textsc{Crane} could be beneficial to
research in combinatorics as well~\cite{DBLP:conf/ilp/BarvinekB0ZK21}.

>>>>

\crefname{algocf}{Algorithm}{Algorithms}
\Crefname{algocf}{Algorithm}{Algorithms}
\crefname{condition}{Condition}{Conditions}
\Crefname{condition}{Condition}{Conditions}
\crefname{line}{line}{lines}

@@

\crefalias{formula}{equation}
\crefname{formula}{Formula}{Formulas}
\Crefname{formula}{Formula}{Formulas}
\creflabelformat{formula}{#2\textup{(#1)}#3}

@@

\makeatletter
\newcommand{\nosemic}{\renewcommand{\@endalgocfline}{\relax}}% Drop semi-colon ;
\newcommand{\dosemic}{\renewcommand{\@endalgocfline}{\algocf@endline}}% Reinstate semi-colon ;
\newcommand{\pushline}{\Indp}% Indent
\newcommand{\popline}{\Indm\dosemic}% Undent
\makeatother

@@

% TODO: (everywhere in the paper): emphasise that the experimental results show
% Crane to be strictly more capable than ForcLift in a sense that it can solve
% everything that ForcLift can and more (including instances beyond C^2).

@@

\emph{First-order model counting} (FOMC) is the problem of computing the number
of models of a sentence in first-order logic given the size(s) of its
domain(s)~\cite{DBLP:conf/pods/BeameBGS15}. \emph{Symmetric weighted FOMC}
(WFOMC) extends FOMC with (pairs of) weights on predicates and asks for a
weighted sum across all models instead. By fixing the sizes of the domains, a
WFOMC instance can be rewritten as an instance of (propositional) weighted model
counting~\cite{DBLP:journals/ai/ChaviraD08}. WFOMC emerged as the dominant
approach to \emph{lifted (probabilistic) inference}. Lifted inference techniques
exploit symmetries in probabilistic models by reasoning about sets rather than
individuals~\cite{DBLP:conf/ecai/Kersting12}. By doing so, many instances become
solvable in polynomial time~\cite{DBLP:conf/nips/Broeck11}. The development of
lifted inference algorithms coincided with work on probabilistic relational
models that combine the syntactic power of first-order logic with probabilistic
models such as Bayesian and Markov networks, allowing for a more relational view
of uncertainty
modelling~\cite{DBLP:series/synthesis/2016Raedt,DBLP:journals/ml/KimmigMG15,DBLP:journals/ml/RichardsonD06}.
Lifted inference techniques for probabilistic databases, while developed
somewhat independently, have also been inspired by
WFOMC~\cite{DBLP:journals/pvldb/GatterbauerS15,DBLP:journals/debu/GribkoffSB14}.
While WFOMC has received more attention in the literature, FOMC is an
interesting problem in an of itself because of its connections to finite model
theory~\cite{DBLP:conf/kr/BremenK21} and applications in enumerative
combinatorics~\cite{DBLP:conf/ilp/BarvinekB0ZK21,DBLP:journals/corr/abs-2302-04606}.
There is also a recent line of work that uses FOMC-based methods to count
solutions to constraint satisfaction
problems~\cite{DBLP:journals/jair/TotisDRK23}.

@@

@@

% However, none of the publicly available exact (W)FOMC algorithms can efficiently
% compute functions as simple as a factorial.\footnote{There is some previous work
%   discussing FOMC for
%   factorials~\cite{DBLP:journals/jair/Kuzelka21,DBLP:conf/kr/BremenK21,DBLP:conf/ijcai/Broeck16}.
%   One of them~\cite{DBLP:conf/ijcai/Broeck16} is an example `by hand' that does
%   not directly correspond to any algorithm.
%   Another~\cite{DBLP:journals/jair/Kuzelka21} proves that the problem is
%   liftable but contains no experimental results. The
%   third~\cite{DBLP:conf/kr/BremenK21} contains experimental results performed
%   with an algorithm that is not publicly available.}
% Two more papers to mention:
% 1. The L2C paper has the factorial example as well. But then why doesn't L2C
% work on it?
% 2. \cite{DBLP:journals/corr/abs-2302-04606} computes n! using publicly
% unavailable code

% TODO: add a different introductory problem statement
We claim that the capabilities of (W)FOMC algorithms can be significantly
extended by empowering them with the ability to construct recursive solutions.
The topic of recursion in the context of WFOMC has been studied before but in
limited ways. Barv{\'{\i}}nek et al.~\shortcite{DBLP:conf/ilp/BarvinekB0ZK21}
use WFOMC to generate numerical data that is then used to conjecture recurrence
relations that explain that data. Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11} introduced the idea of \emph{domain
  recursion}. Intuitively, domain recursion partitions a domain of size $n$ into
a single explicitly named constant and the remaining domain of size $n-1$.
However, many stringent conditions are enforced to ensure that the search for a
tractable solution always terminates.

@@

\begin{definition}[Clause]\label{def:clause}
  A \emph{clause}
  is a triple $c = (L, E, \delta_c)$, where $L$ is a set of literals, $E$ is a
  set of constraints, and $\delta_c$ is the domain map of $c$. Let $\Vars$ be
  the function that maps clauses and sets of either literals or constraints to
  the set of variables in them. In particular,
  $\Vars(c) \coloneqq \Vars(L) \cup \Vars(E)$. \emph{Domain map}
  $\delta_{c}\colon \Vars(c) \to \mathcal{D}$ is a function that maps all
  variables in $c$ to their domains such that (s.t.) if $(X, Y) \in E$ for some
  variables $X$ and $Y$, then $\delta_c(X) = \delta_c(Y)$. For convenience, we
  sometimes write $\delta_c$ for the domain map of $c$ without unpacking $c$
  into its three constituents.
\end{definition}

@@

One can read a formula in our format as a sentence in first-order logic. All
variables in a clause are implicitly universally quantified, and all clauses in
a formula are implicitly linked by a conjunction. Thus, formula $\phi$ from
\cref{example:first} reads as
\begin{equation}\label[formula]{eq:formula}
  \begin{split}
    &\begin{multlined}[t]
      (\forall X \in \Gamma\text{. }\forall Y, Z \in \Delta\text{. }\\
      Y \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(X, Z)) \land{}
    \end{multlined}\\
    &\begin{multlined}[t]
      (\forall X, Z \in \Gamma\text{. }\forall Y \in \Delta\text{. }\\
      X \ne Z \Rightarrow \neg \texttt{p}(X, Y) \lor \neg \texttt{p}(Z, Y)).
    \end{multlined}
  \end{split}
\end{equation}

There are two major differences between
\cref{def:constraint,def:clause,def:formula} and the corresponding concepts
introduced by Van den Broeck et
al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11}.\footnote{Van den Broeck et
  al.~\shortcite{DBLP:conf/ijcai/BroeckTMDR11} refer to clauses and formulas as
  c-clauses and c-theories, respectively.} First, we decouple variable-to-domain
assignments from constraints and move them to a separate function $\delta_{c}$
in \cref{def:clause}. Second, while they allow for equality constraints and
constraints of the form $X \not\in \Delta$ for some variable $X$ and domain
$\Delta$, we exclude such constraints simply because they are not needed. Note
that if, e.g., $Y \ne Z$ in \cref{eq:formula} is replaced by $Y = Z$, then
\cref{eq:formula} can be simplified to have one fewer variables. Similarly, if
the same inequality is replaced by $Y = c$ for some constant $c$, then $Y$ can
be eliminated as well. Since constraints are always interpreted as preconditions
for the disjunction of literals in the clause (as in \cref{eq:formula}),
equality constraints can be eliminated without any loss in expressivity.

@@

\section{First-Order Computational Graphs}\label{sec:methods}

@@

In this paper, we introduce \emph{first-order computational graphs} (FCGs) that
generalise circuits by dispensing with acyclicity. Formally, an FCG is a (weakly
connected) directed graph with a single source, node labels, and ordered
outgoing edges. Node labels consist of two parts: the \emph{type} and the
\emph{parameters}. The type of a node determines its out-degree. We make the
following changes and additions to the node types already supported by
\textsc{ForcLift}. First, we introduce a new node type for constraint removal
($\CR$). Second, we replace domain recursion with generalised domain recursion
($\GDR$). And third, for reasoning about partially-constructed FCGs, we write
$\star$ for a placeholder type that is yet to be replaced. We write $T_p$ for an
FCG that has a node with label $T_p$ (i.e., type $T$ and parameter(s) $p$) and
$\star$'s as all of its direct successors. We also write $T_p(v)$ for an FCG
with one edge from a node labelled $T_{p}$ to some other node $v$ (and no other
nodes or edges). See \cref{fig:examplefcg} for an example FCG\@.

@@

\begin{algorithm}[t]
  \caption{The compilation rule for $\GDR$ nodes.}\label{alg:domainrecursion}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{domain $\Omega \in \mathcal{D}$ s.t.\ there is $c \in \phi$ and $X \in \Vars(L_c)$ s.t. $\delta_c(X) = \Omega$\label{line:condition}}{
    $\phi' \gets \emptyset$\;
    $x \gets \text{a new constant in domain } \Omega$\;\label{line:constant}
    \ForEach{clause $c = (L, E, \delta) \in \phi$\label{line:forclause}}{
      $V \gets \{\, X \in \Vars(L) \mid \delta(X) = \Omega \,\}$\;\label{line:V}
      \ForEach{$W \subseteq V$ s.t. $W^2 \cap E = \emptyset$ {\bf and} $W \cap \{\, X \in \Vars(E) \mid (X, y) \in E \text{ for some constant } y \,\} = \emptyset$\label{line:conditions}}{
        \tcc{$\delta'$ restricts $\delta$ to the new set of variables}
        $\phi' \gets \phi' \cup \{\, (L[x/W], E[x/W] \cup \{\, (X, x) \mid (X \in V \setminus W) \,\}, \delta') \,\}$\;\label{line:generation}
      }
    }
    $S \gets S \cup \{\, (\GDR, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

  \Cref{line:conditions} iterates over all subsets $W \subseteq V$ of variables
  that can be replaced by a constant without resulting in evidently
  unsatisfiable formulas. We impose two restrictions on $W$. First,
  $W^2 \cap E = \emptyset$ ensures that there are no pairs of variables in $W$
  that are constrained to be distinct, since that would result in an $x \ne x$
  constraint after substitution. Similarly, we want to avoid variables in $W$
  that have inequality constraints with constants: after the substitution, such
  constraints would transform into inequality constraints between two constants.
  In this case, both subsets of $V$ satisfy these conditions, and
  \cref{line:generation} generates two clauses for the output formula:
  \begin{multline*}
    (\{\, \neg \texttt{p}(X, Y), \neg \texttt{p}(X, Z) \,\}, \{\, (Y, Z), (X, x) \,\}, \\
    \{\, X \mapsto \Gamma, Y \mapsto \Delta, Z \mapsto \Delta \,\}),
  \end{multline*}
  from $W = \emptyset$ and
  \[
    (\{\, \neg \texttt{p}(x, Y), \neg \texttt{p}(x, Z) \,\}, \{\, (Y, Z) \,\}, \{\, Y \mapsto \Delta, Z \mapsto \Delta \,\})
  \]
  from $W = V$.

@@

Note that \Cref{thm:correctness1} is equivalent to Proposition~3 by Van den
Broeck~\shortcite{DBLP:conf/nips/Broeck11}. For the proofs of this and other
theorems, see supplementary material.

@@

\begin{definition}\label{def:replaceable}
  With respect to a formula $\phi$, a pair $(\Omega, x)$ of a domain
  $\Omega \in \mathcal{D}$ and its element $x \in \Omega$ is called
  \emph{replaceable} if
  \begin{enumerate*}[label=(\roman*)]
    \item $x$ does not occur in any literal of any clause of $\phi$, and
    \item for each clause $c = (L, E, \delta_c) \in \phi$ and variable
    $X \in \Vars(c)$, either $\delta_c(X) \ne \Omega$ or $(X, x) \in E$.
  \end{enumerate*}
\end{definition}

\begin{algorithm}[t]
  \caption{The compilation rule for $\CR$ nodes.}\label{alg:constraintremoval}
  \KwIn{formula $\phi$, set of all relevant domains $\mathcal{D}$}
  \KwOut{set of chips $S$}
  $S \gets \emptyset$\;
  \ForEach{replaceable pair $(\Omega \in \mathcal{D}, x \in \Omega)$\label{line:crconditions}}{
    add a new domain $\Omega'$ to $\mathcal{D}$\;\label{line:newdomain}
    $\phi' \gets \emptyset$\;
    \ForEach{clause $(L, E, \delta) \in \phi$}{
      $E' \gets \{\, (a, b) \in E \mid b \ne x \,\}$\;\label{line:constraintremoval}
      \nosemic$\delta' \gets X \mapsto
      \begin{cases}
        \Omega' & \text{if } \delta(X) = \Omega\\
        \delta(X) & \text{otherwise;}
      \end{cases}$\;\label{line:newdelta}
      $\phi' \gets \phi' \cup \{\, (L, E', \delta') \,\}$\;
    }
    $S \gets S \cup \{\, (\CR_{\Omega \mapsto \Omega'}, \langle\phi'\rangle) \,\}$\;
  }
\end{algorithm}

@@

\begin{theorem}[Correctness of $\CR$]
  Let $\phi$ be the input formula of \cref{alg:constraintremoval}, $(\Omega, x)$
  a replaceable pair, and $\phi'$ the formula constructed by the algorithm, when
  $(\Omega, x)$ is selected on \cref{line:crconditions}. Then
  $\phi \equiv \phi'$, where the domain $\Omega'$ introduced on
  \cref{line:newdomain} is interpreted as $\Omega \setminus \{\, x \,\}$.
\end{theorem}

There is no analogue to $\CR$ in previous work on first-order knowledge
compilation. $\CR$ plays a key role by recognising when the constraints of a
formula essentially reduce the size of a domain by one and extracting this
observation into the definition of a new domain. This then allows us to relate
maps between sets of domains to the arguments of a function call.
\Cref{sec:ref,sec:interpret} describe this process.

@@

\paragraph{Hashing.}
We use (integer-valued) hash functions to discard pairs of formulas that are too
different for recursion to work. The hash code of a clause
$c = (L, E, \delta_{c})$ (denoted by $\# c$) combines the hash codes of the sets
of constants and predicates in $c$, the numbers of positive and negative
literals, the number of inequality constraints $|E|$, and the number of
variables $|\Vars(c)|$. The hash code of a formula $\phi$ combines the hash
codes of all its clauses and is denoted by $\#\phi$.

@@

As \textsc{ForcLift} only supports $\Reff$ nodes for when the two formulas are
equal, our approach is significantly more general and can even construct more
complex recursive calls such as $f(n-k-2)$.

@@

\begin{itemize}
  \item TODO: is maybe `parameter name' (or something similar) would be more
        appropriate than `variable name'?
  \item We call a variable name is \emph{atomic} if it is free of subtraction
        (e.g., $m$, $n$); otherwise it is \emph{non-atomic} (e.g., $m-1$,
        $n-l$).
  \item We assume a fixed total ordering of all domains names. In particular, in
        our running example we assume that $\Gamma$ goes before $\Delta$.
  \item For each node $v$, let $\mathfrak{D}(v)$ denote the (precomputed) list
        of domains, the sizes of which we would use as parameters if we were to
        define a function that begins at $v$. As a set, it is computed by
        iteratively setting
        $\mathfrak{D}(v) \gets \left(\bigcup_{u} \mathfrak{D}(u) \setminus I_{v}\right) \cup U_{v}$
        until convergence, where:
        \begin{enumerate*}[label=(\roman*)]
          \item the union is over all direct successors of $v$,
          \item $I_{v}$ is the set of domains \emph{introduced} at node $v$, and
          \item $U_{v}$ is the set of domains \emph{used} by $v$.
        \end{enumerate*}
        The set is then sorted according to the fixed ordering of domains.
  \item $\mathsf{F}$ is a partial map from the set of FCG nodes to function
        names (e.g., $f$, $g$).
  \item Functions \newVariableName and \newFunctionName generate
        previously-unused names for variables and functions, respectively. The
        latter also records the new name in $\mathsf{F}$.
  \item Note that a node is a direct successor of a $\Reff$ node if and only if
        it has an in-degree greater than one.
  \item For simplicity, we refer to FCG nodes using their types.
  \item The algorithm consists of two main functions: \visit and \actuallyVisit.
        The former handles new function definitions, and the latter produces an
        algebraic interpretation of each node depending on its type. As there
        are many node types, \cref{alg:interpretation} only includes the ones
        pertinent to the contributions of this paper; see previous
        work~\cite{DBLP:conf/ijcai/BroeckTMDR11} for information about the other
        types.
  \item Given some node $v$ as input, both \visit and \actuallyVisit return a
        pair $(\mathsf{e}, \mathsf{funs})$. Here, $\mathsf{e}$ is the algebraic
        expression that should be used to evaluate $v$ or use it in another
        expression, and $\mathsf{funs}$ is a list of auxiliary functions that
        were created while constructing $\mathsf{e}$.
\end{itemize}

@@

However, the most important direction for future work is to fully automate this
new process. First, we need an algorithm that transforms FCGs into definitions
of functions. Formalising this process would also allow us to prove the
correctness of the new compilation rules in constructing FCGs that indeed
compute the right (weighted) model counts. Second, these definitions must be
simplified before they can be used, perhaps by a computer algebra system. Third,
we need a way to find the base cases for the recursive definitions provided by
\textsc{Crane}. Fourth, since the first solution found by \textsc{Crane} is not
always optimal in terms of its complexity, an automated way to determine the
asymptotic complexity of a solution would be helpful as well. Achieving these
goals would enable \textsc{Crane} to automatically constructing efficient ways
to compute various functions and sequences. In addition to the potential impact
on areas of artificial intelligence (AI) such as statistical relational
AI~\cite{DBLP:series/synthesis/2016Raedt}, \textsc{Crane} could be beneficial to
research in combinatorics as
well~\cite{DBLP:conf/ilp/BarvinekB0ZK21,DBLP:journals/corr/abs-2302-04606}.
